Supplementary Appendix

### 2025-07-22 paper/preprint/appendix.tex
finished table and rerendered preprint

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be (1) plausible with
respect to the underlying data and (2) actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals ad-hoc during the
training phase to minimize the divergence between learned
representations and plausible, actionable explanations. We demonstrate
empirically and theoretically that our proposed method facilitates
training models that deliver inherently desirable explanations while
promoting robustness and preserving high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

\section{Introduction}\label{introduction}

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules that are carefully hand-crafted by humans, modern
machine learning (ML) models are tasked with learning representations
directly from data, guided by narrow objectives such as predictive
accuracy (\citeproc{ref-goodfellow2016deep}{Goodfellow, Bengio, and
Courville 2016}). Modern advances in computing have made it possible to
provide such models with ever-growing degrees of freedom to achieve that
task, which frequently allows them to outperform traditionally more
parsimonious models. Unfortunately, in doing so, models learn
increasingly complex and highly sensitive representations that humans
can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
(DL) revolution, Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2014}) showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AEs):
perturbed versions of data instances that yield vastly different model
predictions despite being ``imperceptible'' in that they are
semantically indifferent from their factual counterparts. Even though
some partially effective mitigation strategies have been proposed---most
notably \textbf{adversarial training}
(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
2015})---truly robust deep learning remains unattainable even for models
that are considered ``shallow'' by today's standards
(\citeproc{ref-kolter2023keynote}{Kolter 2023}).

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so-called ``lottery ticket hypothesis'' suggests that modern
neural networks can be pruned by up to 90\% while preserving their
predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
Carbin 2019}). Similarly, Zhang et al.
(\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are expressive enough to fit randomly
labeled data. Thus, looking at the predictive performance alone, the
solutions may seem to provide compelling explanations for the data, when
in fact they are based on purely associative, semantically meaningless
patterns. This poses two challenges. Firstly, there is no dependable way
to verify if representations correspond to meaningful, plausible
explanations. Secondly, even if we could resolve the first challenge, it
remains undecided how to ensure that models can \emph{only} learn
valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models (see,
e.g., (\citeproc{ref-oneil2016weapons}{O'Neil 2016})). Effective XAI
tools can aid us in monitoring models and providing recourse to
individuals to turn negative outcomes (e.g., ``loan application
rejected'') into positive ones (e.g., ``application accepted''). In line
with this, our work builds upon \textbf{counterfactual explanations}
(CE) proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
approach to achieve this goal. CEs prescribe minimal changes for factual
inputs that, if implemented, would prompt some fitted model to produce a
desired output.

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Def. \ref{def-explainability}). Instead, the choice has generally been
to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are in fact
compatible with plausible explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that improved explanatory capacity
can arise as a consequence of regularization techniques aimed at other
training objectives such as generative capacity, generalization, or
robustness (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
\citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
\citeproc{ref-schut2021generating}{Schut et al. 2021}). As further
discussed in Section~\ref{sec-lit}, our work consolidates these findings
within a single objective.

\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
novel training regime explicitly geared towards improving the
explainability of models. In high-level terms, we define this concept as
as the extent to which valid explanations derived for an opaque model
are also deemed plausible with respect to the underlying data and the
global actionability constraints. To the best of our knowledge, our
framework represents the first attempt to address this challenge by
employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we demonstrate that CT substantially improves
explainability and positively contributes to the robustness of trained
models without sacrificing predictive performance. Finally, in
Section~\ref{sec-discussion}, we discuss the challenges and conclude
that CT is a promising approach towards making opaque models more
trustworthy.

\section{Related Literature}\label{sec-lit}

To make the desiderata for our framework more concrete, we follow
Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
explainability to the quality of CEs that can be generated for a given
model. The authors show that CEs---understood as minimal input
perturbations that yield some desired model prediction---tend to be more
meaningful if the underlying model is more robust to adversarial
examples. We can make intuitive sense of this finding when looking at
adversarial training (AT) through the lens of representation learning
with high degrees of freedom. As argued before, learned representations
may be sensitive to producing implausible explanations and mispredicting
for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to
``unlearn'' susceptiblity to such examples, adversarial training can
effectively remove implausible explanations from the solution space.

\subsection{Adversarial Examples are Counterfactual
Explanations}\label{adversarial-examples-are-counterfactual-explanations}

This interpretation of the link between explainability through
counterfactuals on one side and robustness to adversarial examples on
the other is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrates that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argues that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate that
augmenting the training data of image classifiers can improve
generalization. Finally, Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) proposes an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on identifying similar
input samples with different annotations and ensuring that the gradient
of the classifier aligns with the vector between such pairs of
counterfactual inputs using the cosine distance as the loss function.

In the natural language processing (NLP) domain, CEs have also been used
to improve models through data augmentation. Wu et al.
(\citeproc{ref-wu2021polyjuice2}{2021}) proposes \emph{Polyjuice}, a
general-purpose counterfactual generator for language models. The
authors empirically demonstrate that the augmentation of training data
with their method improves robustness in a number of NLP tasks.
Balashankar et al. (\citeproc{ref-balashankar2023improving}{2023})
similarly uses \emph{Polyjuice} to augment NLP datasets through diverse
counterfactuals and show that classifier robustness improves by up to
20\%. Finally, Luu and Inoue
(\citeproc{ref-luu2023counterfactual}{2023}) introduces Counterfactual
Adversarial Training (CAT) that also aims to improve generalization and
robustness of language models through a three-step procedure: the
authors identify training samples that are subject to high predictive
uncertainty, generate CEs for them, and fine-tune the language model on
a dataset augmented with the CEs.

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples. Pointing
to clear similarities in how CEs and AEs are generated, Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) makes the case for
jointly studying the opaqueness and robustness problems in
representation learning. Formally, AEs can be seen as the subset of CEs
for which misclassification is achieved
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) shows
that CEs and AEs are equivalent under certain conditions.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of the post-hoc counterfactual explanations. Firstly,
Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
proposes a way to train models that guarantee individual recourse to
some positive target class with high probability. Their approach builds
on adversarial training by explicitly inducing susceptibility to
targeted adversarial examples for the positive class. Additionally, the
proposed method allows for imposing a set of actionability constraints
ex-ante. For example, users can specify that certain features are
immutable. Secondly, Guo, Nguyen, and Yadav
(\citeproc{ref-guo2023counternet}{2023}) is the first to propose an
end-to-end training pipeline that includes CEs as part of the training
procedure. In particular, they propose a specific network architecture
that includes a predictor and CE generator network, where the parameters
of the CE generator network are learnable. Counterfactuals are generated
during each training iteration and fed back to the predictor network. In
contrast to Guo, Nguyen, and Yadav
(\citeproc{ref-guo2023counternet}{2023}), we impose no restrictions on
the ANN architecture at all.

\subsection{Aligning Representations with Plausible
Explanations}\label{aligning-representations-with-plausible-explanations}

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) shows that explainability
can be improved through model averaging and refined model objectives.
The authors propose a way to generate counterfactuals that are maximally
faithful to the model in that they are consistent with what the model
has learned about the underlying data. Formally, they rely on tools from
energy-based modelling (\citeproc{ref-teh2003energy}{Teh et al. 2003})
to minimize the divergence between the distribution of counterfactuals
and the conditional posterior over inputs learned by the model. Their
proposed counterfactual explainer, \emph{ECCCo}, yields plausible
explanations if and only if the underlying model has learned
representations that align with them. The authors find that both deep
ensembles (\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan,
Pritzel, and Blundell 2017}) and joint energy-based models (JEMs)
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data
(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
mitigates the aforementioned risk of relying on a single locally optimal
representations that corresponds to semantically meaningless
explanations for the data. Previous work of Schut et al.
(\citeproc{ref-schut2021generating}{2021}) similarly found that
generating plausible (``interpretable'') CEs is almost trivial for deep
ensembles that have also undergone adversarial training. The case for
JEMs is even clearer: they involve a hybrid objective that induces both
high predictive performance and generative capacity
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This is
closely related to the idea of aligning models with plausible
explanations and has inspired our CT objective.

\section{Counterfactual Training}\label{sec-method}

In this section we propose our novel counterfactual training objective.
In CT, we combine ideas from adversarial training, counterfactual
explanations, and energy-based modelling with the explicit goal of
aligning representations with plausible explanations that comply with
user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which counterfactuals comply with the
underlying data-generating process
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020}). Plausibility is
a necessary but insufficient condition for using CEs to provide
algorithmic recourse (AR) to individuals (negatively) affected by opaque
models. To be actionable, AR recommendations must also be attainable. A
plausible CE for a rejected 20-year-old loan applicant, for example,
might reveal that their application would have been accepted, if only
they were 20 years older. Ignoring all other features, this would comply
with the definition of plausibility if 40-year-old individuals were in
fact more credit-worthy on average than young adults. But of course this
CE does not qualify for providing actionable recourse to the applicant
since \emph{age} is not a (directly) mutable feature. Counterfactual
training aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata: plausibility and
actionability. Formally, we define explainability as follows:

\begin{definition}[Model Explainability]
\label{def-explainability}
Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.

We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:

\begin{enumerate}
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
\end{enumerate}
where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
\end{definition}

\noindent The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
Intuitively, plausible counterfactuals are consistent with the data and
faithful counterfactuals are consistent with what the model has learned
about input data. Actionability constraints in Def.
\ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we focus on domain and
mutability constraints for individual features \(x_d\) for
\(d=1,...,D\). We limit ourselves to classification tasks for reasons
discussed in Section~\ref{sec-discussion}.

\subsection{Our Proposed Objective}\label{our-proposed-objective}

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
let \(T\) vary and consider the counterfactual search as converged as
soon as the predicted probability for the target class has reached a
pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see supplementary appendix.}

To train models with high explainability as defined in Def.
\ref{def-explainability}, we propose to leverage counterfactuals in the
following objective:

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and, (2) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-off
between the components can governed through penalties
\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

\subsection{Directly Inducing Explainability with Contrastive
Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}

As observed by Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}), any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimizes the
contrastive divergence between the observed samples from
\(p(\mathbf{x})\) and generated samples from \(p_\theta(\mathbf{x})\).

A key empirical finding in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
well with respect to the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) relies on
Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative
prior for initialization but we depart from their methodology. Instead
of SGLD, we propose to use counterfactual explainers to generate
counterfactuals of observed training samples. Specifically, we have:

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while increasing the
energy of counterfactuals (negative samples)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the
counterfactuals get more plausible (Def. \ref{def-explainability})
during training, these opposing effects gradually balance each other out
(\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).

The departure from SGLD of (\citeproc{ref-grathwohl2020your}{Grathwohl
et al. 2020}) allows us to tap into the vast repertoire of explainers
that have been proposed in the literature to meet different desiderata.
For example, many methods facilitate the imposition of domain and
mutability constraints. In principle, any existing approach for
generating counterfactual explanations is viable, so long as it does not
violate the faithfulness condition. Like JEMs
(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), CT can be
considered a form of contrastive representation learning.

\subsection{Indirectly Inducing Explainability with Adversarial
Robustness}\label{indirectly-inducing-explainability-with-adversarial-robustness}

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
(\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023};
\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or adversarial
examples (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
only sensible choice, is inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2014}) that defines an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) that considers
misclassification as the key distinguishing feature of AE. One of the
key observations of this work is that we can leverage CEs during
training and get adversarial examples essentially for free, which can be
used to reap the aforementioned benefits of adversarial training.

\subsection{Encoding Actionability Constraints}\label{sec-constraints}

>>>>

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\usepackage{siunitx}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Supplementary Appendix},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training }
\title{Supplementary Appendix}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning
